Red Wine Analysis with R By Christopher Boone
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages and surpress messages.

library(ggplot2)
library(knitr)
library(dplyr)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
myData <- read.csv(file='wineQualityReds.csv')

```

The wineQualityReds data set contains 12 variables that describe different 
aspects of 1599 different red wines. The variables consist of: fixed acidity,
volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide,
total sulfur dioxide, density, pH, sulphates, alcohol, and quality.

# Univariate Plots Section


```{r echo=FALSE, message=FALSE, Univariate_Plots}

qplot(x = quality, data = myData, bins = 10,
      main = 'Count of Red Wine by Quality',
      xlab = 'Quality',
      ylab = 'Count') +
  stat_count(breaks = 3:8)

```

We can see by this graph that the wine quality has an even distribution and 
is centralized  on 5 and 6 which gives us an approx median of 6.






```{r echo=FALSE}
str(myData)
myData$quality.level <- ifelse(myData$quality < 5, "Below Average", 
                             ifelse(myData$quality < 7, 
                                    "Average", "Above Average"))
myData$quality.level <- factor(myData$quality.level, 
                             levels=c("Above Average", 
                                      "Average", "Below Average"), ordered=TRUE)
attach(myData)

```


```{r echo=FALSE}

qplot(x = alcohol, data = myData, bins = 25,
      main = 'Count of Red Wine by Alcohol Content',
      xlab = 'Alcohol Content',
      ylab = 'Count')

```

When we tak a count of wines by alcohol content we can see that the count is 
right skewed. We can also see that we have some outliers in the data set.

```{r echo=FALSE}

qplot(x = residual.sugar, data = myData, bins = 25,
      main = 'Count of Red Wine by Residual Sugar Content',
      xlab = 'Risidual Sugar',
      ylab = 'Count')

median(residual.sugar)

mean(residual.sugar)

```

We can see that residual sugar is right skewed and have a few 
groups of outliers. The median sits at 2.2 and a mean of 2.54.

```{r echo=FALSE}

qplot(x = sulphates, data = myData, bins = 25,
      main = 'Count of Red Wine by Sulphate Levels',
      xlab = 'Sulphates',
      ylab = 'Count')

median(sulphates)

mean(sulphates)


```


We can see that Sulphates are Right skewed with a median of 0.62 and
mean of 0.658


```{r echo=FALSE}

qplot(x = free.sulfur.dioxide, data = myData, bins = 75,
      main = 'Count of Red Wine by Free Sulfur Dioxide Amounts',
      xlab = 'Free Sulfur Dioxide Amounts',
      ylab = 'Count')

median(free.sulfur.dioxide)

mean(free.sulfur.dioxide)

```

With sulfur Dioxide counts we have a right skewed graph with median of 14
and mean of 15.87.


```{r echo=FALSE}

qplot(x = total.sulfur.dioxide, data = myData, bins = 25,
      main = 'Count of Red Wine by Total Sulfur Dioxide Amounts',
      xlab = 'Total Sulfur Dioxide Amounts',
      ylab = 'Count')

median(total.sulfur.dioxide)

mean(total.sulfur.dioxide)

```

We can see that Total Sulfur Dioxide amounts are right skewed with some far 
outliers. The median is 38 and mean is 46.47.


```{r echo=FALSE}

qplot(x = density, data = myData, bins = 20,
      main = 'Count of Red Wine by Density',
      xlab = 'Density',
      ylab = 'Count')

median(density)

mean(density)

```

We can see that Density has a normal distribution with a median of 0.99675 and 
mean of 0.9967467.


```{r echo=FALSE}

qplot(x = pH, data = myData, bins = 40,
      main = 'Count of Red Wine by pH Levels',
      xlab = 'pH',
      ylab = 'Count')

median(pH)

mean(pH)

```

We can see that pH levels have a normal distribution with a median of 3.31 and
mean of 3.311113.


```{r echo=FALSE}

qplot(x = citric.acid, data = myData, bins = 40,
      main = 'Count of Red Wine by Citric Acid Content',
      xlab = 'Citric Acid',
      ylab = 'Count')

```

We can see that Citric Acid Content has a non-normal distribution.


```{r echo=FALSE}

qplot(x = volatile.acidity, data = myData, bins = 40,
      main = 'Count of Red Wine by Volitile Acidity',
      xlab = 'Volatile Acidity',
      ylab = 'Count')

median(volatile.acidity)

mean(volatile.acidity)

```

The distribution of Volatile Acidity seems to have a right skew to it that 
has a median of 0.52 and mean of 0.528.

```{r echo=FALSE}

qplot(x = fixed.acidity, data = myData, bins = 40 ,
      main = 'Count of Red Wine by Fixed Acidity',
      xlab = 'Fixed Acidity',
      ylab = 'Count')

a = median(fixed.acidity)

b = mean(fixed.acidity)

```

The distribution of Fixed Acidity has a right skew to it that 
has a median of 7.9 and mean of 8.319637.

# Univariate Analysis

### What is the structure of your dataset?
This dataset contains 12 variables on 1599 different red wines. 

We can also see that:

  Median quality of 5.5
  
  Most of the variables are right skewed.
  
  The bulk of the wines pH levels range from 3 to 3.6
  
  There are some outliers that exist that may need to be ommitted 
      in further analysis.

### What is/are the main feature(s) of interest in your dataset?
I am mainly focusing on the quality as compared pH levels. 
I want to see if there is a coorrelation between how the quality
of wine is rated and the various acidity types.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The acidity levels will also play a factor in this data set due to that fact 
that they directly impact the pH levels.

### Did you create any new variables from existing variables in the dataset?

I created groupings of quality based on the normal distribution of the count of
wines by quality. This led me to create three groups which consists of:

Above Average > 7
Average < 7 > 5
Below Average   < 5

By doing this it will help me to categorize the sets and make it easier to see
any differences that may lie between the win qualities.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

As I used an existing dataset that was already clean I did not perform any 
cleaning operations. One graph that did stand out was the Citric Acid which had
a non-normal distribution.


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}

ggplot(aes(x = quality.level, y = alcohol), data = myData) +
  geom_boxplot()

```

This plot shows that the higher rated wines tend to have a higher alcohol 
content. It is interesting to see that there is a slight dip in alcohol content
when you move from the below average to average groups of wine.

```{r echo=FALSE}

ggplot(aes(x = quality.level, y = volatile.acidity), data = myData) +
  geom_boxplot()

```

From this graph we see that there is an obvious relationship between the quality
of a wine and the volitile acidity level. Lower the volatile acidity,
higher the quality of wine.

```{r echo=FALSE}

ggplot(aes(x = quality.level, y = fixed.acidity), data = myData) +
  geom_boxplot()

```

From this graph we see that there is an obvious relationship between the quality
of a wine and the fixed acidity level. Higher the fixed acidity, 
higher the quality of wine.


```{r echo=FALSE}

ggplot(aes(x = quality.level, y = citric.acid), data = myData) +
  geom_boxplot()

```

From this graph we see that there is an obvious relationship between the quality
of a wine and the citric acid level. Higher the citric acid, 
higher the quality of wine.


```{r echo=FALSE}

qplot(x = alcohol, density, data = myData,
      xlab = 'Density',
      ylab = 'Alcohol')

```

We can see by this plot that the density tends to be lower for higher quality
wines though the range is much larger than the other groupings.

```{r echo=FALSE}

qplot(x = fixed.acidity, citric.acid, data = myData,
      xlab = 'Fixed Acidity',
      ylab = 'Citric Acid')

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I observed:

  The relationship of citric acid to quality is negative.
  The relationship of fixed acidity to quality is negative.
  The relationship of volitile acidity to quality is positive.
  
I found it interesting that there was a slight dip in the alcohol content range
for the average grouping when compared to the other tow groups. It is also
interesting to see that the volitive and fixed acidity levels when compared to
quality level are inverse of each other.

This shows us that people tend to enjoy windes that have a higher level of 
fixed acidity and citric acid while maintaining lower levels 
of volatile acidity.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Some of the interesting things I observed were:

  The relationship of alcohol to density is negative.

  THe relationship of fixed acidity to citric acid is positive.

### What was the strongest relationship you found?

I found that Quality is most strongly associated with higher fixed 
acidity as well as lower volatile acidity.

By using this relationship I believe it would be possible to accuratly determine
what quality level a wine would be rated from these characteristics.

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

ggplot(aes(x = fixed.acidity, y = volatile.acidity),
       data = subset(myData, !is.na(quality.level))) +
      geom_point(aes(color = quality.level)) 
      
```

We can see here that lower quality wines tend to have higher volatile acidity
while having a lower fixed acidity. There are a few outliers in the below average
range that violate this.

We can also see that above average wines tend to have a volitile acidity no more
than .8. Seeing this I can tell that volitile acidity has a much more 
significant affect on wine quality than fixed acidity does.


```{r echo=FALSE}


ggplot(data = myData,
       aes(x = citric.acid, y = fixed.acidity,
                      color = quality.level)) +
  geom_point() +
  facet_wrap(quality.level)

      
```

Here we can see that across the three groups that as fixed acidity and citric 
acid increase, the quality level decreases.

```{r echo=FALSE}


ggplot(data = myData,
       aes(x = volatile.acidity, y = fixed.acidity,
                      color = quality.level)) +
  geom_point() +
  facet_wrap(quality.level)
      
```

From here we can see that as the quality of the wine increases, the higher the
fixed acidity and lower the volatile acidity become.+

```{r echo=FALSE}


ggplot(data = myData,
       aes(x = volatile.acidity, y = citric.acid,
                      color = quality.level)) +
  geom_point() +
  facet_wrap(quality.level)
      
```
This graph shows that citric acid doesn't always have a significant impact on
the quality of the wine.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

When comparing quality level, fixed acidity, and citric acid together and 
splitting it into their respective groups it became easier to see how the 
presence of high fixed acidity and citric acid are indicators of a wines 
quality.

The other thing that stood out to me was the relationship of volatile acidity,
fixed acidity, and quality went together. The graph helped to make clear that 
fixed acidity has less of an impact on wine quality than volatile does.

### Were there any interesting or surprising interactions between features?

Overall, I found that comparing what I believe to be the key indicators to rate
a quality wine that citric acid had less of a determining factor. Yes, you can 
see that as the overall quality of wine increases, the citric acid content 
gets higher but the final graph did a good job of showing the overlap with 
citric acid that exists between the three groups.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}


ggplot(data = myData,
       aes(x = citric.acid, y = fixed.acidity,
                      color = quality.level)) +
  geom_point() +
  labs(x = 'Fixed Acidity', y = 'Citric Acid') +
  ggtitle('Plot One: Fixed Acidity vs Citric Acid by Quality Level') +
  facet_wrap(quality.level) 

      
```

### Description One

I chose this graph because I think it does the best job at showing how 
citric acid and fixed acidity are two of the more influential factors when
it comes to the quality of a wine.

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(aes(x = quality.level, y = citric.acid, fill = quality.level),
       data = myData, 
       ) +
  geom_boxplot() +
  labs(x = 'Quality Level', y = 'Citric Acid', fill = 'Quality Level') +
  ggtitle('Plot 2: Citric Acid vs Quality Level')

```

### Description Two

Here we can see that citric acid is proven to have a positive relationship with
the quality of wine. It is interesting to see that the range of what is 
considered an above average wine in relation to citric acid content
is much smaller than the average and below average groups. I also find it 
intregueing that there is a lot of overlap between the average and 
above average groups.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x = quality.level, y = fixed.acidity, fill = quality.level),
       data = myData) +
  geom_boxplot() +
  labs(x = 'Quality Level', y = 'Fixed Acidity', fill = 'Quality Level') +
  ggtitle('Plot 2: Fixed Acidity vs Quality Level')
```

### Description Three

Lastly, I selected this graph because it shows the other acid that has one of
the greater impacts on the quality of wine. Using the second and third plot
you can better understand what you are seeing in the first plot and gain a 
deeper understanding of what the information is showing.
------

# Reflection

WThe red wine dataset contained information on 1599 different red wines and 
12 different attributes on each. As I preformed my exploration I tried to find
different ways to visualize the relationship of the various acid types and how 
it affects the quality of the wine. I discovered that some styles of plots are
much better at presenting less specific data where as others are best for 
comparing single variables. For instance, I had to break the jitter plot into
the three seperate quality groups because if I kept them overlapped it makes it
difficult to pull any meaning from the plot. The use of single variable plots
and graphs allows you to gain more granular insight on specific attributes. 
By doing this I can better understand the behavior of the dataset and identify
any anamolies that may make interpreting a more complex, multivariable graph
difficult.

Moving forward, I believe growing this dataset to the point that it looks at
the geographic location of the people who rate the wines would be a great 
benefit. With that, we could identify each country and the preferred attributes
that they like in their wine. By doing this the wine makers would be able to
start targeting different countries and allocate wines with certain 
characteristics to areas that best prefer them.

### Struggles / Successes

Throughout the project I spent a lot of time inputing different graph types for
the same metrics over and over to see which did best to represent the 
information I was looking at. I struggle for a bit in trying to find a graph 
that was easy to read without getting too confusing with points on top of 
points. I ended up discovering how to apply facet_wrap to point and jitter 
plots which allowed me to reduce the clutter making the plots more readable 
and easy to use.
